\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=0.8in, top=0.6in}
\usepackage{times}
\usepackage{enumitem}
\setlist{nosep}
\usepackage[compact]{titlesec}
\titleformat*{\section}{\large\bfseries}
\titleformat*{\subsection}{\normalsize\bfseries}
\usepackage{amsmath}
\usepackage{amssymb}

\title{\large\bfseries 4D-MLLM: Learning a 4D Latent World Model from 2D Video \\ via Structured Spatiotemporal Transformers}
\author{Aiden Li}
\date{}

\begin{document}

\maketitle

\section{Introduction}
\label{sec:introduction}

The current landscape in video language modeling is paradoxical. Video-LLMs can generate fluent descriptions and answer many semantic questions from RGB video \cite{zhang2023videollamainstructiontunedaudiovisuallanguage, cheng2024videollama2advancingspatialtemporal, zhang2024llavanextvideo}. However, when questions require stable 3D structure over time, such as tracking objects through occlusion, comparing motion under viewpoint changes, or maintaining correct temporal causality, performance degrades sharply. This gap is now quantified by spatiotemporal benchmarks, which reveal substantial deficits in temporal coherence, viewpoint-aware motion understanding, and physical plausibility \cite{zhou2025vlm4dspatiotemporalawarenessvision, zhu20254dbenchbenchmarkingmultimodallarge, wang2025spatial4dbenchversatile4dspatial}.

These trends point to a central conclusion: the bottleneck is not a missing module but the backbone representation itself. This proposal targets the missing piece: a backbone that learns a unified 4D latent world representation from the most available modality, timestamped 2D video. We introduce 4D-MLLM, a latent world model transformer designed around three key changes: (1) redefining tokens from 2D patches to 4D entities, (2) modifying attention to be biased by spatiotemporal proximity, and (3) shifting the representation target from video features to a unified 4D latent world representation. These changes aim to bridge semantic video understanding and structured 4D (3D+time) reasoning using ubiquitous 2D video inputs.

\section{Related Work}
\label{sec:related}

\subsection{Sequence-based Reasoning}
Foundational Video-LLMs like Video-LLaMA 2 \cite{cheng2024videollama2advancingspatialtemporal} and LLaVA-NeXT-Video \cite{zhang2024llavanextvideo} have established high semantic benchmarks by processing video as 2D patch sequences. However, these models often suffer from ``temporal flickering'' and lack physical grounding, as evidenced by deficits in benchmarks like VLM4D \cite{zhou2025vlm4dspatiotemporalawarenessvision} and Spatial4D-Bench \cite{wang2025spatial4dbenchversatile4dspatial}. Recent developments, such as the introduction of Spatio-Temporal Rotary Positional Embeddings (ST-RoPE) in Qwen2.5-VL and Gemini 2.0, attempt to bridge this ``signal-to-structure'' gap. Despite these advances, the newer DynPhys-Bench \cite{zhang2025abenchphysicsbenchmarkingphysicalreasoning} indicates that sequence-based models still struggle with non-linear motion and occlusions. This has led to the emergence of frameworks like DATE \cite{yuan2025date}, which introduce absolute chronometric awareness through Timestamp Injection Mechanisms (TIM) to ground visual tokens within a precise temporal coordinate system, suggesting that a transition to continuous 4D manifold representations is necessary.

\subsection{Native 4D Perception}
Emerging works have moved beyond simple 4D point cloud alignment toward Neural Field-based architectures. Early native 4D systems like 4DPC²hat \cite{zhang20264dpc2hatdynamicpointcloud} and B4DL \cite{Choi_2025} ingest dynamic point cloud or LiDAR sequences to align 4D geometry with language. This has evolved into more sophisticated frameworks like Uni4D-LLM \cite{zhou2025uni4dllmunifiedspatiotemporalawarevlm}, which unifies 4D understanding and generation by extracting semantic and geometric features simultaneously through adaptive cross-attention. Recent models, such as 4DLangVGGT \cite{wu20254dlangvggt4dlanguagevisualgeometry} and Lyra \cite{bahmani2025lyra}, utilize 4D Gaussian Splatting (4D-GS) as a mid-level representation. Unlike prior methods requiring explicit 3D supervision, current Gaussian Splatting-based LLM (GS-LLM) architectures leverage Differentiable Rendering Layers and self-distillation from video diffusion models to backpropagate language-based grounding losses directly into latent 4D scene graphs \cite{wang2025vggtvisualgeometrygrounded}, bridging raw pixels and structured geometry.

\subsection{Latent World Models}
A parallel direction involves injecting geometric priors or learning intrinsic world models. Geometry-augmented VLMs like DSR Suite \cite{zhou2025learningreason4ddynamic} and PSG-4D-LLM \cite{yang20244dpanopticscenegraph} utilize reconstruction-derived tokens or 4D panoptic scene graphs to enhance spatial reasoning. This trend has recently led to the emergence of Latent World Models (LWM) such as NVIDIA Cosmos \cite{nvidia2025worldsimulationvideofoundation}. These models demonstrate that massive video pre-training on billions of frames can lead to ``emergent 4D geometry,'' where consistent 3D views can be synthesized from 2D inputs. 4D-MLLM builds on this by formalizing these emergent properties into a backbone paradigm that ensures representations are both visually plausible and geometrically grounded.

\section{Proposed Method}
\label{sec:methodology}

We propose a backbone that transforms the fundamental unit of computation from a ``2D patch'' to a ``4D token''. For each token, we maintain a tuple $T_i = (f_i, x_i, \tau_i)$, where $f_i$ denotes appearance features, $x_i$ is a learned latent 3D coordinate that organizes tokens into a geometry-consistent space, and $\tau_i$ is the absolute timestamp associated with the observation. Importantly, $x_i$ does not correspond to ground-truth geometry and does not require explicit 3D reconstruction supervision; it is learned as part of the representation target so that the backbone naturally supports questions of ``what is where, when'' and can maintain object permanence across views and time.

To learn $x_i$ from timestamped 2D video, we induce geometry through self-supervised spatiotemporal consistency, rather than relying on SLAM-style reconstruction pipelines. We train the model to align tokens that correspond to the same physical entity across time and viewpoint changes, enforcing stability for static regions ($x_{i,t} \approx x_{j,t+1}$) while allowing structured evolution for dynamic regions. Time is modeled as a continuous variable using learnable Fourier features to handle variable frame rates and precise durations. Importantly, the model learns absolute time understanding by encoding real-world timestamps directly in the representation, which enables consistent reasoning over long videos, irregular sampling, and multi-camera synchronization. We represent dynamics with a time-conditioned predictor that relates changes in $x$ to elapsed time, encouraging temporal scaling consistency.

On top of these 4D tokens, we modify the attention mechanism to reflect spatiotemporal structure: attention is biased so that tokens that are close in latent 3D space and close in time interact more strongly than distant ones. To manage long-horizon consistency, we implement a dual-memory system inspired by 3DLLM-Mem \cite{hu20253dllmmemlongtermspatialtemporalmemory}, utilizing a Memory Fusion Module to selectively attend to relevant features from an episodic 3D memory bank. The backbone is trained with Masked 4D Modeling by masking a spatial region (in latent space) or a temporal interval and requiring reconstruction from the remaining context, which encourages both occlusion reasoning and trajectory completion. We evaluate 4D-MLLM as a backbone rather than a task-specific head, using benchmarks that expose 4D reasoning gaps including VLM4D \cite{zhou2025vlm4dspatiotemporalawarenessvision}, 4D-Bench \cite{zhu20254dbenchbenchmarkingmultimodallarge}, and Spatial4D-Bench \cite{wang2025spatial4dbenchversatile4dspatial}, with comparisons to strong video-language baselines \cite{zhang2023videollamainstructiontunedaudiovisuallanguage, zhang2024llavanextvideo} and ablations over token definition, attention bias, and representation objectives.

\section{Expected Outcomes}
\label{sec:outcomes}

We expect 4D-MLLM to establish a backbone architecture that learns stable, geometry-consistent spatiotemporal representations from timestamped 2D video. By redefining tokens as latent 4D entities and incorporating time-aware structure into attention and training objectives, the model should improve viewpoint consistency, object permanence under occlusion, and long-horizon temporal reasoning. These improvements are expected to yield measurable gains on benchmarks that stress 3D awareness, motion understanding, and temporal causality, while also demonstrating that structured 4D world representations can emerge through self-supervised learning. More broadly, this work aims to shift video-language modeling from sequence-level semantic encoding toward physically grounded, temporally coherent world modeling suitable for downstream reasoning and embodied applications.

\newpage
\bibliographystyle{plain}
\bibliography{references}

\end{document}
